Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2024-07-28 Thesis/03-chap3.qmd
some sections should be second level headers instead of first level
update according to emi edited

Plotting residuals is commonly regarded as a standard practice in linear regression diagnostics [see @belsley1980regression; @cook1982residuals]. This visual assessment plays a crucial role in identifying deviations from model assumptions, such as linearity, homoscedasticity, and normality. It also helps in understanding the goodness of fit and various characteristics of the model.

Generating a residual plot in most statistical software is often as straightforward as executing a line of code or clicking a button. However, accurately interpreting a residual plot can be challenging. A residual plot can exhibit various visual features, but it is crucial to recognize that some may arise from the characteristics of predictors and the inherent randomness of the error, rather than indicating a violation of model assumptions [@li2023plot]. Consider @fig-false-finding as an example, the residuals display a triangular shape pointing to the left. While this might suggest heteroskedasticity, it is important to avoid over-interpreting this visual pattern. In this case, the fitted regression model is correctly specified, and the triangular shape is actually a result of the skewed distribution of the predictors, rather than indicating a flaw in the model.

@@

However, as pointed out by @li2023plot, a primary limitation of the lineup protocol lies in its reliance on human judgments. Unlike conventional statistical tests that can be performed computationally in statistical software, the lineup protocol requires human evaluation of images. This characteristic makes it less suitable for large-scale applications, given the associated high labour costs and time requirements.

There is a substantial need to develop an approach that alleviates analysts' workload by automating repetitive tasks and providing standardized results in a controlled environment. The large-scale evaluation of lineups is impractical without the use of technology and machines.

The utilization of computers to interpret data plots has a rich history, with early efforts such as "Scagnostics" by @tukey1985computer, focusing on scatter plot diagnostics. @wilkinson2005graph expanded on this work, introducing graph theoretic scagnostics, which encompassed computable measures applied to planar proximity graphs. These measures, including, but not limited to, "Outlying", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy", and "Striated" aimed to characterize outliers, shape, density, trend, coherence and other characteristics of the data. While this approach has been inspiring, there is a recognition [@buja2009statistical] that it may not capture all the necessary visual features that differentiate true residual plots from null plots. A more promising alternative entails enabling machines to learn the function for extracting visual features from residual plots. Essentially, this means empowering computers to discern the crucial visual features for residual diagnostics and determining the method to extract them.

Modern computer vision models are well-suited for addressing this challenge. They rely on deep neural networks with convolutional layers [@fukushima1982neocognitron]. These layers leverage hierarchical patterns in data, downsizing and transforming images by summarizing information in a small space. Numerous studies have demonstrated the efficacy of convolutional layers in addressing various vision tasks, including image recognition [@rawat2017deep]. Despite the widespread use of computer vision models in fields like computer-aided diagnosis [@lee2015image], pedestrian detection [@brunetti2018computer], and facial recognition [@emami2012facial], their application in reading data plots remains limited. While some studies have explored the use of computer vision models for tasks such as reading recurrence plots for time series regression [@ojeda2020multivariate], time series classification [@chu2019automatic; @hailesilassie2019financial; @hatami2018classification; @zhang2020encoding], anomaly detection [@chen2020convolutional], and pairwise causality analysis [@singh2017deep], the application of reading residual plots with computer vision models represents a relatively new field of study.

In this paper, we develop computer vision models and integrate them into the residual plots diagnostics workflow, addressing the need for automated visual inference. The paper is structured as follows. @sec-model-specifications discusses various specifications of the computer vision models. @sec-model-distance-between-residual-plots defines the distance measure used to measure model violations. @sec-model-distance-estimation explains how the computer vision models estimate this distance measure. @sec-model-statistical-testing covers the statistical testing based on the estimated distance. Sections [-@sec-model-data-generation], [-@sec-model-architecture], and [-@sec-model-training] detail the data preparation, model architecture, and training process, respectively. The results are presented in @sec-model-results. Finally, we conclude with a discussion of our findings and propose ideas for future research directions.

@@

Deep learning models are in general very sensitive to the input data. The quality and relevance of the input data greatly influence the model's capacity to generate insightful and meaningful results. There are several designs of the input format can be considered.

A straightforward design involves feeding a vector of residuals along with a vector of fitted values, essentially providing all the necessary information for creating a residuals vs fitted values plot. However, a drawback of this method is the dynamic input size, which changes based on the number of observations. For modern computer vision models implemented in mainstream software like TensorFlow [@abadi2016tensorflow], the input shape is typically fixed. One solution is to pad the input vectors with leading or trailing zeros when the input tensor expects longer vectors, but it may fail if the input vector surpasses the designed length. Another strategy is to summarize the residuals and fitted values separately using histograms and utilize the counts as the input. By controlling the number of bins in the histograms, it becomes possible to provide fixed-length input vectors. Still, since histograms only capture the marginal distribution of residuals and fitted values respectively, they can not be used to differentiate visual patterns with same marginal distributions but different joint distributions.

Another design involves using an image as input. The primary advantage of this design, as opposed to the vector format, is the availability of the existing and sophisticated image processing architectures developed over the years, such as the VGG16 architecture proposed in @simonyan2014very. These architectures can effectively capture and summarize spatial information from nearby pixels, which is less straightforward with vector input. The main considerations are the image resolution and the aesthetics of the residual plot. In general, higher resolution provides more information to the model but comes with the trade-off of increased complexity and greater difficulty in training. As for the aesthetics of the residual plot, a practical solution is to consistently present residual plots in the same style to the model. This implies that the model can not accept arbitrary images as input but requires the use of the same pre-processing pipeline to convert residuals and fitted values into a standardized-style residual plot.

@@

We did not explore all the mentioned input formats due to the significant costs tied to data preparation and model training. Taking into account the implementation cost and the need for model interpretability, we settled on the single residual plot input format.

@@

Given that the input is a single residual plot represented as a fixed-resolution image, the output from the computer vision model can take one of two forms: binary or numeric. This choice determines whether the model belongs to a classification model or a regression model. The binary outcome encoded as $0$ and $1$ could be used to represent whether the input image is a null plot, or whether the input image would be rejected in a visual test conducted by humans. Training a model following the latter option requires data from prior human subject experiments, presenting difficulties in controlling the quality of data due to variations in experimental settings across different studies. Additionally, some visual inference experiments are unrelated to linear regression models or residual plot diagnostics, resulting in a limited amount of available training data.

@@

In this study, we chose to define and use a distance measure to quantify the difference between the residual plot and the null plots. @vo2016localizing have also demonstrated that defining a proper distance between images can enhance the matching accuracy in image search compared to a binary outcome model.

## Distance Between Residual Plots {#sec-model-distance-between-residual-plots}

In a visual test, the observer will be asked to choose one or more plots that stand out as most distinct from others in a given lineup. To develop a computer vision model for assessing residual plots within the visual inference framework, it is important to precisely define a numerical measure of "difference" or "distance" between plots. This distance can take the form of a basic statistical operation on pixels, such as the sum of square differences. Alternatively, it could involve established image similarity metrics like the Structural Similarity Index Measure [@wang2004image]. The challenge lies in the fact that metrics tailored for image comparison may not be suitable for evaluating data plots, where only essential plot elements require assessment [@chowdhury2018measuring]. Furthermore, scagnostics mentioned in @sec-model-introduction could be used to construct distance measures for residual plots comparison, but the functional form still needs to be carefully refined to accurately reflect the extent of the violations.

@@

The distance measure proposed in this study takes into account the fact that we tried to measure how different a residual plot is from a good residual plot, or in other words, how different a given fitted regression model is from a correctly specified model.

For a classical normal linear regression model, residuals $\boldsymbol{e}$ are derived from the fitted values $\hat{\boldsymbol{y}}$ and observed values $\boldsymbol{y}$. Suppose the data generating process is known and the regression model is correctly specified, by the Frisch-Waugh-Lowell theorem [@frisch1933partial], residuals $\boldsymbol{e}$ can also be treated as random variables and written as a linear transformation of the error $\boldsymbol{\varepsilon}$ formulated as $\boldsymbol{e} = \boldsymbol{R}\boldsymbol{\varepsilon}$, where $\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'$ is the residual operator and an idempotent matrix, $\boldsymbol{X}$ is the design matrix, $\boldsymbol{I}_n$ is a $n$ by $n$ identity matrix, and $n$ is the number of observations.

One of the assumptions of the classical normal linear regression model is the error $\boldsymbol{\varepsilon}$ follows a multivariate normal distribution with zero mean and constant variance, i.e., $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)$. It can be known that residuals $\boldsymbol{e}$ also follow a certain probability distribution transformed from the multivariate normal distribution, which will be denoted as $Q$. This reference distribution $Q$ summarizes what good residuals should follow given the design matrix $\boldsymbol{X}$ is known and fixed.

Suppose the design matrix $\boldsymbol{X}$ has linearly independent columns, the trace of the hat matrix will equal to the number of columns in $\boldsymbol{X}$ denoted as $k$. As a result, the rank of $\boldsymbol{R}$ is $n - k$, and $Q$ is a degenerate multivariate distribution. To capture the characteristics of $Q$, such as moments, we can simulate a large numbers of $\boldsymbol{\varepsilon}$ and transform it to $\boldsymbol{e}$ to get the empirical estimates. For simplicity, in this study, we replaced the variance-covariance matrix of residuals $\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}' = \boldsymbol{R}\sigma^2$ with a full-rank diagonal matrix $\text{diag}(\boldsymbol{R}\sigma^2)$, where $\text{diag}(.)$ sets the non-diagonal entries of a matrix to zeros. The resulting distribution for $\boldsymbol{Q}$ is $N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))$.

@@

This distance measure was first proposed in @li2023plot. It was mainly designed for measuring the effect size of non-linearity and heteroskedasticity in a residual plot. @li2023plot have derived that, for a classical normal linear regression model that omits a necessary higher-order predictors $\boldsymbol{Z}$, and incorrectly assumes $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)$ while in fact $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})$ with $\boldsymbol{V}$ being an arbitrary symmetric positive semi-definite matrix, $Q$ can be represented as $N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))$. Note that the variance-covariance matrix is replaced with the diagonal matrix to ensure it is a full-rank matrix.

@@

In order to approximate $D_{KL}$ in a practically computable manner, the elements of $\boldsymbol{e}$ are assumed to be independent of each other. This assumption solves both of the issues mentioned above. First, we no longer need to integrate over $n$ random variables. The result of @eq-kl-1 is now the sum of the Kullback-Leibler divergence evaluated for each individual residual thanks for the independence assumption. Second, it is not required to know the joint probability density $p(\boldsymbol{e})$ any more. Instead, the evaluation of Kullback-Leibler divergence for an individual residual relies on the knowledge of the marginal density $p_i(e_i)$, where $e_i$ is the $i$-th residual for $i = 1, ..., n$. This is much easier to approximate through simulation. It is also worth mentioning that this independence assumption generally will not hold, since $\text{cov}(e_i, e_j) \neq 0$ if $\boldsymbol{R}_{ij} \neq 0$ for any $1 \leq i < j \leq n$, but its existence is essential for reducing the computational cost.

@@

We tried to train a computer vision model to estimate this distance measure with a residual plot. Let $D$ be the result of @eq-kl-0, and our estimator $\hat{D}$ is formulated as

@@

\noindent where $V_{h \times w}(.)$ is a plotting function that saves a residuals vs fitted values plot with fixed aesthetic as an RGB image with $h \times w$ pixels, $f_{CV}(.)$ is a computer vision model which takes an $h \times w$ image as input and predicts the distance in the domain $[0, +\infty)$.

With the estimated distance $\hat{D}$, we will be able to know how different the underlying distribution of the residuals is from a good residual distribution. $\hat{D}$ can also be used as an index of the model violations indicating the strength of the visual signal embedded in the residual plot.

@@

## Model Violations Index (MVI) {#sec:model-violations-index}

$\hat{D}$ is an estimator of the difference between the true residual distribution and the reference residual distribution. This difference primarily arises from deviations in model assumptions. The magnitude of $D$ directly reflects the degree of these deviations, thus making $\hat{D}$ instrumental in forming a model violations index.

@@

```{r}
#| label: factor
#| tbl-cap: "Factors used in the data generating process for synthetic data simulation. Factor $j$ and $a$ controls the non-linearity shape and the heteroskedasticity shape respectively. Factor $b$, $\\sigma_\\varepsilon$ and $n$ control the signal strength. Factor $\\text{dist}_\\varepsilon$, $\\text{dist}_{x1}$ and $\\text{dist}_{x2}$ specifies the distribution of $\\varepsilon$, $X_1$ and $X_2$ respectively."
tibble(Factor = c("j", "a", "b", "$\\beta_1$", "$\\beta_2$", " $\\text{dist}_\\varepsilon$", "$\\text{dist}_{x1}$", "$\\text{dist}_{x2}$", "$\\sigma_{\\varepsilon}$", "$\\sigma_{X1}$", "$\\sigma_{X2}$", "n"),
       Domain = c("\\{2, 3, ..., 18\\}", "[-1, 1]", "[0, 100]", "\\{0, 1\\}", "\\{0, 1\\}", "\\{discrete, uniform, normal, lognormal\\}", "\\{discrete, uniform, normal, lognormal\\}", "\\{discrete, uniform, normal, lognormal\\}", "[0.0625, 9]", "[0.3, 0.6]", "[0.3, 0.6]", "[50, 500]")) %>%
  kableExtra::kable(escape = FALSE,
                    booktabs = TRUE)
```

@@

# Limitations and Future Work

@@

# Conclusion

In this study, we have introduced a distance measure based on Kullback-Leibler divergence to quantify the disparity between the residual distribution of a fitted classical normal linear regression model and the reference residual distribution assumed under correct model specification. This distance measure effectively captures the magnitude of model violations in misspecified models. We propose a computer vision model to estimate this distance, utilizing the residual plot of the fitted model as input. The resulting estimated distance serves as the foundation for constructing a single Model Violation Index (MVI), facilitating the quantification of various model violations.

>>>>

Plotting residuals is commonly regarded as a standard practice in linear regression diagnostics [@belsley1980regression; @cook1982residuals]. This visual assessment plays a crucial role in identifying deviations from model assumptions, such as linearity, homoscedasticity, and normality. It also helps in understanding the goodness of fit and various unexpected characteristics of the model.

Generating a residual plot in most statistical software is often as straightforward as executing a line of code or clicking a button. However, accurately interpreting a residual plot can be challenging. A residual plot can exhibit various visual features, but it is crucial to recognize that some may arise from the characteristics of predictors and the natural stochastic variation of the observational unit, rather than indicating a violation of model assumptions [@li2023plot]. Consider @fig-false-finding as an example, the residual plot displays a triangular left-pointing shape. The distinct difference in the spread of the residuals across the fitted values may result in the analyst suggesting that there may be heteroskedasticity, however, it is important to avoid over-interpreting this visual pattern. In this case, the fitted regression model is correctly specified, and the triangular shape is actually a result of the skewed distribution of the predictors, rather than indicating a flaw in the model.

@@

A primary limitation of the lineup protocol lies in its reliance on human judgements [see @li2023plot about the practical limitations]. Unlike conventional statistical tests that can be performed computationally in statistical software, the lineup protocol requires human evaluation of images. This characteristic makes it less suitable for large-scale applications, given the associated high labour costs and time requirements. There is a substantial need to develop an approach to substitute these human judgement with an automated reading of data plots using machines.

The utilization of computers to interpret data plots has a rich history, with early efforts such as "Scagnostics" by @tukey1985computer, a set of numerical statistics that summarise features of scatter plots. @wilkinson2005graph expanded on this work, introducing scagnostics based on computable measures applied to planar proximity graphs. These measures, including, but not limited to, "Outlying", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy", and "Striated" aimed to characterize outliers, shape, density, trend, coherence and other characteristics of the data. While this approach has been inspiring, there is a recognition [@buja2009statistical] that it may not capture all the necessary visual features that differentiate true residual plots from null plots. A more promising alternative entails enabling machines to learn the function for extracting visual features from residual plots. Essentially, this means empowering computers to discern the crucial visual features for residual diagnostics and determining the method to extract them.

Modern computer vision models are well-suited for addressing this challenge. They rely on deep neural networks with convolutional layers [@fukushima1982neocognitron]. These layers leverage hierarchical patterns in data, downsizing and transforming images by summarizing information in a small space. Numerous studies have demonstrated the efficacy of convolutional layers in addressing various vision tasks, including image recognition [@rawat2017deep]. Despite the widespread use of computer vision models in fields like computer-aided diagnosis [@lee2015image], pedestrian detection [@brunetti2018computer], and facial recognition [@emami2012facial], their application in reading data plots remains limited. While some studies have explored the use of computer vision models for tasks such as reading recurrence plots for time series regression [@ojeda2020multivariate], time series classification [@chu2019automatic; @hailesilassie2019financial; @hatami2018classification; @zhang2020encoding], anomaly detection [@chen2020convolutional], and pairwise causality analysis [@singh2017deep], the application of reading residual plots with computer vision models is a new field of study.

In this paper, we develop computer vision models and integrate them into the residual plots diagnostics workflow, addressing the need for an automated visual inference. The paper is structured as follows. @sec-model-specifications discusses various specifications of the computer vision models. @sec-model-distance-between-residual-plots defines the distance measure used to measure model violations. @sec-model-distance-estimation explains how the computer vision models estimate this distance measure. @sec-model-statistical-testing covers the statistical testing based on the estimated distance. Sections [-@sec-model-data-generation], [-@sec-model-architecture], and [-@sec-model-training] detail the data preparation, model architecture, and training process, respectively. The results are presented in @sec-model-results. Finally, we conclude with a discussion of our findings and propose ideas for future research directions.

@@

Deep learning models are in general very sensitive to the input data. The quality and relevance of the input data greatly influence the model's capacity to generate insightful and meaningful results. There are several designs of the input format that can be considered.

A straightforward design involves feeding a vector of residuals along with a vector of fitted values, essentially providing all the necessary information for creating a residuals vs fitted values plot. However, a drawback of this method is the dynamic input size, which changes based on the number of observations. For modern computer vision models implemented in mainstream software like TensorFlow [@abadi2016tensorflow], the input shape is typically fixed. One solution is to pad the input vectors with leading or trailing zeros when the input tensor expects longer vectors, but it may fail if the input vector surpasses the designed length.

Another strategy is to summarize the residuals and fitted values separately using histograms and utilize the counts as the input. By controlling the number of bins in the histograms, it becomes possible to provide fixed-length input vectors. Still, since histograms only capture the marginal distribution of residuals and fitted values respectively, they can not be used to differentiate visual patterns with same marginal distributions but different joint distributions.

Another design involves using an image as input. The primary advantage of this design, as opposed to the vector format, is the availability of the existing and sophisticated image processing architectures developed over the years, such as the VGG16 architecture proposed in @simonyan2014very. These architectures can effectively capture and summarize spatial information from nearby pixels, which is less straightforward with vector input. The main considerations are the image resolution and the aesthetics of the residual plot. In general, a higher resolution provides more information to the model but comes with a trade-off of increased complexity and greater difficulty in training. As for the aesthetics of the residual plot, a practical solution is to consistently present residual plots in the same style to the model. This implies that the model can not accept arbitrary images as input but requires the use of the same pre-processing pipeline to convert residuals and fitted values into a standardized-style residual plot.

@@

Taking into account the implementation cost and the need for model interpretability, we used the single residual plot input format in this paper.

@@

Given that the input is a single residual plot represented as a fixed-resolution image, we can choose the output from the computer vision model to be either binary (classification) or numeric (regression).

The binary outcome can represent whether the input image is consistent with a null plot as determined by either (1) the data generating process or (2) the result of a visual test based on human judgement. Training a model following the latter option requires data from prior human subject experiments, presenting difficulties in controlling the quality of data due to variations in experimental settings across different studies. Additionally, some visual inference experiments are unrelated to linear regression models or residual plot diagnostics, resulting in a limited amount of available training data.

@@

In this study, we chose to define and use a distance between a true residual plot and a theoretical "good" residual plot. @vo2016localizing have also demonstrated that defining a proper distance between images can enhance the matching accuracy in image search compared to a binary outcome model.

## Distance from a Theoretical "Good" Residual Plot  {#sec-model-distance-between-residual-plots}

To develop a computer vision model for assessing residual plots within the visual inference framework, it is important to precisely define a numerical measure of "difference" or "distance" between plots. This distance can take the form of a basic statistical operation on pixels, such as the sum of square differences. Alternatively, it could involve established image similarity metrics like the Structural Similarity Index Measure [@wang2004image]. The challenge lies in the fact that metrics tailored for image comparison may not be suitable for evaluating data plots, where only essential plot elements require assessment [@chowdhury2018measuring]. Furthermore, scagnostics mentioned in @sec-model-introduction could be used to construct distance measures for residual plots comparison, but the functional form still needs to be carefully refined to accurately reflect the extent of the violations.

In this section, we will introduce a distance measure between a true residual plot and a theoretical 'good' residual plot. This measure quantifies the divergence between the residual distribution of a given fitted regression model and that of a correctly specified model. The computation assumes knowledge of the data generating processes for predictors and response variables. Since these processes are often unknown in practice, we will discuss a method to estimate this distance using a computer vision model in Section @sec-model-distance-estimation.

@@

For a classical normal linear regression model, $\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\beta} + \boldsymbol{e}$, the residual $\hat{\boldsymbol{e}}$ are derived as the difference of the fitted values and observed values $\boldsymbol{y}$. Suppose the data generating process is known and the regression model is correctly specified, by the Frisch-Waugh-Lowell theorem [@frisch1933partial], residuals $\hat{\boldsymbol{e}}$ can also be treated as random variables and written as a linear transformation of the error $\boldsymbol{e}$ formulated as $\hat{\boldsymbol{e}} = \boldsymbol{R}\boldsymbol{e}$, where $\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top$ is the residual operator, $\boldsymbol{I}_n$ is a $n$ by $n$ identity matrix, and $n$ is the number of observations.

One of the assumptions of the classical normal linear regression model is that the error $\boldsymbol{e}$ follows a multivariate normal distribution with zero mean and constant variance, i.e., $\boldsymbol{e} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)$. It can be known that residuals $\hat{\boldsymbol{e}}$ also follow a certain probability distribution transformed from the multivariate normal distribution, which will be denoted as $Q$. This reference distribution $Q$ summarizes what "good" residuals should follow given the design matrix $\boldsymbol{X}$ is known and fixed.

Suppose the design matrix $\boldsymbol{X}$ has linearly independent columns, the trace of the hat matrix will equal to the number of columns in $\boldsymbol{X}$ denoted as $k$. As a result, the rank of $\boldsymbol{R}$ is $n - k$, and $Q$ is a degenerate multivariate distribution. To capture the characteristics of $Q$, such as moments, we can simulate a large numbers of $\boldsymbol{\varepsilon}$ and transform it to $\boldsymbol{e}$ to get the empirical estimates. For simplicity, in this study, we replaced the variance-covariance matrix of residuals $\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}' = \boldsymbol{R}\sigma^2$ with a full-rank diagonal matrix $\text{diag}(\boldsymbol{R}\sigma^2)$, where $\text{diag}(.)$ sets the non-diagonal entries of a matrix to zeros. The resulting distribution for $Q$ is $N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))$.

@@

This distance measure was first proposed in @li2023plot. It was mainly designed for measuring the effect size of non-linearity and heteroskedasticity in a residual plot. @li2023plot have derived that, for a classical normal linear regression model that omits necessary higher-order predictors $\boldsymbol{Z}$ and the corresponding parameter $\boldsymbol{\beta}_z$, and incorrectly assumes $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)$ while in fact $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})$ where $\boldsymbol{V}$ is an arbitrary symmetric positive semi-definite matrix, $Q$ can be represented as $N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))$. Note that the variance-covariance matrix is replaced with the diagonal matrix to ensure it is a full-rank matrix.

@@

In order to approximate $D_{KL}$ in a practically computable manner, the elements of $\boldsymbol{e}$ are assumed to be independent of each other. This assumption solves both of the issues mentioned above. First, we no longer need to integrate over $n$ random variables. The result of @eq-kl-1 is now the sum of the Kullback-Leibler divergence evaluated for each individual residual due to the assumption of independence between observations. Second, it is not required to know the joint probability density $p(\boldsymbol{e})$ any more. Instead, the evaluation of Kullback-Leibler divergence for an individual residual relies on the knowledge of the marginal density $p_i(e_i)$, where $e_i$ is the $i$-th residual for $i = 1, ..., n$. This is much easier to approximate through simulation. It is also worth mentioning that this independence assumption generally will not hold, since $\text{cov}(e_i, e_j) \neq 0$ if $\boldsymbol{R}_{ij} \neq 0$ for any $1 \leq i < j \leq n$, but its existence is essential for reducing the computational cost.

@@

We use a computer vision model to estimate this distance measure with a residual plot. Let $D$ be the result of @eq-kl-0, and our estimator $\hat{D}$ is formulated as

@@

\noindent where $V_{h \times w}(.)$ is a plotting function that saves a residuals vs fitted values plot with fixed aesthetic as an image with $h \times w$ pixels in three channels (RGB), $f_{CV}(.)$ is a computer vision model which takes an $h \times w$ image as input and predicts the distance in the domain $[0, +\infty)$.

With the estimated distance $\hat{D}$, we will be able to know how different the underlying distribution of the residuals is from a theoretical "good" residual distribution. $\hat{D}$ can also be used as an index of the model violations indicating the strength of the visual signal embedded in the residual plot.

@@

## Model Violations Index {#sec:model-violations-index}

$\hat{D}$ is an estimator of the difference between the true residual distribution and the reference residual distribution. This difference primarily arises from deviations in model assumptions. The magnitude of $D$ directly reflects the degree of these deviations, thus making $\hat{D}$ instrumental in forming a model violations index (MVI).

@@

```{r}
#| label: tbl-factor
#| tbl-cap: "Factors used in the data generating process for synthetic data simulation. Factor $j$ and $a$ controls the non-linearity shape and the heteroskedasticity shape respectively. Factor $b$, $\\sigma_\\varepsilon$ and $n$ control the signal strength. Factor $\\text{dist}_\\varepsilon$, $\\text{dist}_{x1}$ and $\\text{dist}_{x2}$ specifies the distribution of $\\varepsilon$, $X_1$ and $X_2$ respectively."
tibble(Factor = c("j", "a", "b", "$\\beta_1$", "$\\beta_2$", " $\\text{dist}_\\varepsilon$", "$\\text{dist}_{x1}$", "$\\text{dist}_{x2}$", "$\\sigma_{\\varepsilon}$", "$\\sigma_{X1}$", "$\\sigma_{X2}$", "n"),
       Domain = c("\\{2, 3, ..., 18\\}", "[-1, 1]", "[0, 100]", "\\{0, 1\\}", "\\{0, 1\\}", "\\{discrete, uniform, normal, lognormal\\}", "\\{discrete, uniform, normal, lognormal\\}", "\\{discrete, uniform, normal, lognormal\\}", "[0.0625, 9]", "[0.3, 0.6]", "[0.3, 0.6]", "[50, 500]")) %>%
  kableExtra::kable(escape = FALSE,
                    booktabs = TRUE)
```

@@

## Limitations and Future Work

@@

## Conclusion

In this paper, we have introduced a distance measure based on Kullback-Leibler divergence to quantify the disparity between the residual distribution of a fitted classical normal linear regression model and the reference residual distribution assumed under correct model specification. This distance measure effectively captures the magnitude of model violations in misspecified models. We propose a computer vision model to estimate this distance, utilizing the residual plot of the fitted model as input. The resulting estimated distance serves as the foundation for constructing a single Model Violation Index (MVI), facilitating the quantification of various model violations.
